---
title: "A one-way ANOVA tutorial"
format:
  arxiv-pdf:
    keep-tex: false 
    linenumbers: false
    doublespacing: false    
    runninghead: "A one-way ANOVA tutorial"
    include-in-header: A-one-way-ANOVA-tutorial-header.tex
  arxiv-html: default
author:
  - name: Dirk Ostwald
    affiliations:
      - name: Otto-von-Guericke-Universität Magdeburg
        department: Institute of Psychology
        address: Universitätsplatz 2
        # city: Magdeburg
        # country: Germany
        # postal-code: 39106
    # orcid: 0000-0002-3530-1845
    # email: dirk.ostwald@ovgu.de
bibliography: A-one-way-ANOVA-tutorial.bib  
---

# Introduction {#sec-intro}

Analysis of variance (ANOVA) is an umbrella term for a large variety of interwoven 
experimental designs and data-analytical methods that aim to quantify the effects
of interventions. As the name implies, ANOVA in its traditional sense is concerned 
with analyzing data variability, seeking to partition it into variation attributable 
to experimental manipulations and variation arising from unsystematic, non-interventional 
sources. ANOVA was popularized a century ago by Ronald A. Fisher as an 
inference method to assess the evidence for an effect of experimental intervention when 
comparing a number of treatment groups (@fisher1925a). Since then, experimental designs have 
diversified tremendously and along with them the data-analytical approaches under
the ANOVA label. The form of ANOVA we are concerned with in the current tutorial
is typically referred to as *one-way ANOVA* and applies in the context of a single
experimental factor, such as different and unrelated treatment groups. In the following, 
we use the term ANOVA to exclusively refer to the one-way ANOVA scenario. In 
@sec-anova-as-a-computational-procedure, we discuss the traditional 
computation-oriented approach to ANOVA. 

Over the course of the 20th century, ANOVA has become regarded as a special 
linear model that describes how data in an ANOVA experimental design is generated. 
This perspective enables experimental effects to be expressed through model parameters 
while also providing a framework for quantifying the uncertainty associated with 
the resulting parameter inferences. While a comprehensive introduction of this 
(general) linear model approach to statistical inference is beyond the scope of this tutorial, we
discuss some of its aspects in @sec-anova-as-a-linear-model.

Finally, over the course of the last forty years, the digital revolution has allowed
the development of computerized algorithms that evaluate parameter estimates numerically
within seconds. This in turn has allowed the development of ANOVA models that are
more flexibly adapted to specific experimental contexts and can directly capture
structural context information of no interest to the evaluation of experimental effects.
Broadly speaking, these methods come under the label of *linear mixed models* 
and dominate the applied use of ANOVA approaches in today's research. In 
@sec-anova-as-a-linear-mixed-model we aim to provide some intuition about these 
approaches in the context of randomized block and nested subsampling designs. 

Throughout, we assume familiarity with the sum symbol (for a brief refresher, 
see @sec-the-sum-symbol). We eschew introducing the matrix formulation of 
linear models, which in fact provides the formal language in which such models 
are most precisely expressed. For the essential probabilistic concepts, we mainly appeal
to intuition. However, please note that a proper understanding of modern ANOVA
can only be achieved based on firm foundations in linear algebra and probability theory.

## An example scenario {#sec-an-example-scenario}

As a first concrete data analysis scenario, we consider the effect of four different 
*treatments* that are assessed based on 10 *observational units* each. The 
observational units that are exposed to one of the treatments will also be referred to 
as a *(treatment) group*. For the example, we assume $p = 4$ treatment groups of 
$m = 10$ observational units per group and thus $n = p\cdot m = 4 \cdot 10 = 40$ 
observational units with associated univariate measurements in total. In the 
language of (factorial) experimental designs the treatments form what is called 
an *experimental factor* and the different variations that the treatment 
can take are referred to as *levels* of the group factor. We label the treatments 
generically as `T1`, `T2`, `T3`, `T4`. 

For a biological example, the observational units could be fibroblasts from a 
given cell culture strain, say `S1` and the measurements could be the amount of 
collagen fiber produced by each fibroblast within 24 hours measured by some 
optical method. The treatments could correspond to different cytokines that may 
or may not increase the production  of collagen fiber production. @tbl-anova-data-set-1 
shows a set of exemplary measurements. 


```{r, echo = F, eval = F}
library(MASS)                                                                   # multivariate normal distribution
set.seed(2)                                                                     # reproducibility 
m       = 10                                                                    # number of observational units per group
p       = 4                                                                     # number of groups 
n       = p*m                                                                   # number of observation units 
Xt      = cbind(                                                                # design matrix
          matrix(1,nrow = n, ncol = 1),                                         # design matrix
          kronecker(diag(p), matrix(1,nrow = m,ncol = 1)))                      # design matrix
X      = Xt[,-2]                                                                # design matrix
I_n    = diag(n)                                                                # n x n identity matrix
beta   = list(matrix(c(5,5,0,-2), nrow = p), matrix(c(5,0,0,0), nrow = p))      # \beta = (\mu_0,\alpha_2,\alpha_3,\alpha_4)
sigsqr = 3                                                                      # \sigma^2   
for(i in 1:2){                                                                  # data set iterations
  y    = mvrnorm(1, X %*% beta[[i]], sigsqr*I_n)                                # 1 realization of an n-dimensional random vectors
  y    = round(y, digits = 1)                                                   # one decimal place 
  D    = data.frame(Treatment = rep(paste0("T", 1:4), each = m), Collagen = y)  # dataframe                   
  write.csv(D,  paste0("./_data/anova-data-set-",i,".csv"), row.names = FALSE)} # save 
```
```{r echo = F, message = F}
#| label: tbl-anova-data-set-1
#| tbl-cap : "Example data set 1"
fname   = "./_data/anova-data-set-1.csv"                                        # filename
D       = read.table(file.path(fname), sep = ",", header = TRUE)                # data loading
D       = data.frame(split(D$Collagen, D$Treatment))                            # wide format
knitr::kable(D, digits = 2, align = "r")                                        # visualization
```

The data set of @tbl-anova-data-set-1 is summarized in terms of its treatment 
group means and treatment group standard deviations and visualized in @fig-anova-data-set-1 
using the **R** code below. Note that all individual data points are shown on 
the figure. The visualization uses the Base R Graphics functionality and does not
require additional packages for full control of all figure properties. The different
functions can be studied using `?functioname` in **R** or with the help of one's 
favorite general purpose large language model, such as ChatGPT.

\tiny
```{r, echo = T, eval = F}
# data loading and group mean and standard deviation evaluation
D       = read.csv("./_data/anova-data-set-1.csv")                              # dataframe in long format                                                
D       = data.frame(split(D$Collagen, D$Treatment))                            # wide format organization                                                            
gmeans  = colMeans(D)                                                           # group means                                                    
gstds   = apply(D,2,sd)                                                         # group standard deviations  

# figure formatting
pdf(                                                                            # figure file format
file      = "./_figures/anova-data-set-1.pdf",                                  # figure file path and name
width     = 7,                                                                  # figure width in inches
height    = 5)                                                                  # figure height in inches
par(                                                                            # figure parameter  
mfcol     = c(1,1),                                                             # 1 x 1 subplot layout
family    = "sans",                                                             # fontfamily (sans serif)
pty       = "m",                                                                # plot area (maximal not square)    
bty       = "l",                                                                # axes in L form        
mgp       = c(2,1,0),                                                           # margin spacing  
font.main = 1,                                                                  # title font
lwd       = 1,                                                                  # linewidth    
las       = 1)                                                                  # orientiation of x-ticks

# bars
x         = barplot(                                                            # barplot and barlocations
gmeans,                                                                         # group mean bars
ylim      = c(0,16),                                                            # y-axis limits
col       = "gray90",                                                           # bar color
ylab      = "Collagen",                                                         # y-axis label
xlab      = "Treatment",                                                        # x-axis label  
main      = "Data Set 1")                                                       # title

# standard deviation error bars
arrows(                                                                         # errorbars                               
x0        = x,                                                                  # x-ordinate
y0        = gmeans - gstds,                                                     # y-ordinate, lower end
x1        = x,                                                                  # x-ordinate  
y1        = gmeans + gstds,                                                     # y-ordinate upper end
code      = 3,                                                                  # caps on both ends
angle     = 90,                                                                 # cap                          
length    = 0.05)                                                               # cap length 
 
# data points
for(i in 1:ncol(D)){                                                            # data point iterations
points(                                                                         # point plot
jitter(rep(x[i], length(D[,i])), amount = .3),                                  # left/right jitter for overlap avoidance
D[,i],                                                                          # data points of interest
pch     = 21,                                                                   # point type
col     = "White",                                                              # outline color  
bg      = "Black")}                                                             # fill color  
dev.off()                                                                       # pdf closing  
```
\normalsize

![Barplot visualization of data set 1. Each dot represents the collagen production
of a single fibroblast undergoing one of the treatment levels](./_figures/anova-data-set-1.pdf){#fig-anova-data-set-1 fig-align="center" width=70%}

As is evident from @fig-anova-data-set-1, Treatment 1 (`T1`) and Treatment 3 (`T3`) induce
similar average collagen production per fibroblast. Treatment 2 (`T2`) induces a 
marked increase collagen production and Treatment 4 (`T4`) induces less collagen 
production.

For comparison purposes, we also consider a second example data set generated 
using fibroblasts from a second cell culture strain `S2` but otherwise identical
to the first example data set. We tabulate this data set in @tbl-anova-data-set-2 and 
visualize it in @fig-anova-data-set-2. Obviously, in this data set, the treatment levels
do not induce differences in the average collagen production per fibroblast.

```{r echo = F, message = F}
#| label: tbl-anova-data-set-2
#| tbl-cap : "Example data set 2"
fname   = "./_data/anova-data-set-2.csv"                                        # filename
D       = read.csv("./_data/anova-data-set-1.csv")                              # dataframe in long format                                                
D       = data.frame(split(D$Collagen, D$Treatment))                            # wide format organization  
knitr::kable(D, digits = 2, align = "r")                                        # visualization
```
![Barplot visualization of data set 2. Each dot represents the collagen production
of a single fibroblast undergoing one of the treatment levels](./_figures/anova-data-set-2.pdf){#fig-anova-data-set-2 fig-align="center" width=70%}

# ANOVA as a computational procedure {#sec-anova-as-a-computational-procedure}

**Empirical variance.** A simple measure of the variability of $n$ scalar data points 
$x_1,...,x_n$ is their *empirical variance*
$$
s^2 = \frac{1}{n-1}\sum_{i=1}^n \left(x_i - \bar{x}\right)^2,
$$ {#eq-empirical-variance}
where $\bar{x} := \frac{1}{n}\sum_{i=1}^n x_i$ denotes the *mean* of the data points. 
Conceptually, $s^2$ in @eq-empirical-variance is computed by first computing
the mean $\bar{x}$ of all data points as a measure of their central tendency. 
Next, the difference $x_i - \bar{x}$ between each data point $x_i$ and the 
the mean $\bar{x}$ of all data points is evaluated. If the data point is larger than 
the mean, this yields a positive number. If the data point is smaller than the 
mean, this yields a negative number. If the data point and the mean are virtually
identical, this yields a number close to zero. To summarize the $n$ deviations 
from the mean into a single number, the differences $x_i - \bar{x}$ are added up. However, 
simply summing the differences $x_i - \bar{x}$ for $i=1,...,n$ will have the effect 
that positive and negative deviations from the mean cancel each other out. 
To prevent this, each difference is squared. The square of any number, positive or 
negative, is always non-negative. Squaring the differences, i.e., evaluating 
$(x_i - \bar{x})^2$ for $i = 1,...n$ thus has the effect of making all terms in the sum of 
@eq-empirical-variance positive. It addition, large deviations between a data
point and the mean contribute a lot to the sum, whereas small deviations contribute 
little. For example, a diffence $x_i - \bar{x} = -5$, after squaring, contributes 
$25$, whereas a difference $x_i-\bar{x} = 0.5$, after squaring contributes only $0.25$. 
Evaluation of the sum of squared deviations then yields a single number that represents 
the variability of the data points about their mean. In @eq-empirical-variance, 
this number is finally standardized by division by $n-1$, i.e., the variability 
is expressed roughly as an average per data point. It would be the exact average,
if instead of dividing by $n-1$, the division would be by $n$. However, for a 
large number of data points, this difference is virtually negligible, as e.g., 
for $n = 100$, $\frac{1}{100} = 0.01$ and $\frac{1}{100-1} = 0.0099\approx 0.01$. 
The motivation of the denominator $n-1$ is best understood in the context 
of designing *unbiased estimators*, a concept which is beyond the scope of this 
tutorial. In general, expressions such as 
$$
\sum_{i=1}^n \left(x_i - \bar{x}\right)^2
$$
in @eq-empirical-variance are called *sum-of-squares*, as they obviously represent 
sums of squared values. Similarly, expressions such as 
$$
\frac{1}{n-1}\sum_{i=1}^n \left(x_i - \bar{x}\right)^2
$$
are called *mean sum-of-squares* as they represent averages of sum of squares.


**Between-sum-of-squares and within-sum-of-squares.** The fundamental idea of ANOVA 
as a computational procedure is to use sum-of-squares to quantify the variability 
observed in an ANOVA data set. To this end, the data variability is partitioned 
in a way that allows for inferring how much of it is likely due to the experimental 
treatment factor and how much of it is due to unsystematic variation. 
The approach is rather simplistic. On the one hand, ANOVA evaluates the variability 
of the individual group means around the grand mean across all observational 
units in all groups as a measure of potential treatment-induced effects. On the 
other hand, ANOVA evaluates the average variability of individual data points 
about their group mean across groups as a measure of unsystematic variation. Note that
the latter variability cannot be treatment-induced, as each data point in a given 
group was exposed to exactly the same treatment level (if the experiment was performed
adequately). Both measures of variability are single numbers and are referred 
to as *between-sum-of-squares* (variability between groups) and *within-sum-of-quares* 
(variability within groups), respectively. Finally, asking how much bigger
or smaller the mean between-sum-of-squares is than the mean within-sum-of-squares 
by dividing the former by the latter yields a single number that indicates the size
of the experimental treatment-induced variability in units of unsystematic variability. 
This number is referred to as the *$F$-value* or *$F$-value* in commemoration of 
Ronald A. Fisher, who famously described this computational procedure in his book 
*Statistical Methods for Research Workers* a century ago (@fisher1925a). 

To represent this logic formally, it is first necessary to introduce some general
notation for the data of a one-way ANOVA design. We here use the convention, that
$y_{ij}$ denotes the $j$th of $n_i$ data points in the $i$th of $p$ groups. Using
this convention, the example data sets of @tbl-anova-data-set-1 and @tbl-anova-data-set-2 can 
be denoted as shown in @tbl-y-ij-notation.

\newpage

| T1        | T2        | T3        | T4        |
|:----------|:----------|:----------|:----------|
| $y_{11}$  | $y_{21}$  | $y_{31}$  | $y_{41}$  |
| $y_{12}$  | $y_{22}$  | $y_{32}$  | $y_{42}$  |
| $y_{13}$  | $y_{23}$  | $y_{33}$  | $y_{43}$  |
| $y_{14}$  | $y_{24}$  | $y_{34}$  | $y_{44}$  |
| $y_{15}$  | $y_{25}$  | $y_{35}$  | $y_{45}$  |
| $y_{16}$  | $y_{26}$  | $y_{36}$  | $y_{46}$  |
| $y_{17}$  | $y_{27}$  | $y_{37}$  | $y_{47}$  |
| $y_{18}$  | $y_{28}$  | $y_{38}$  | $y_{48}$  |
| $y_{19}$  | $y_{29}$  | $y_{39}$  | $y_{49}$  |
| $y_{110}$ | $y_{210}$ | $y_{310}$ | $y_{410}$ |

: One-way ANOVA notation. In this scenario $p = 4$, $n_1 = n_2 = n_3 = n_4 = 10$ and $n = \sum_{i=1}^{p} n_i = 40$ {#tbl-y-ij-notation}

The *between-sum-of-square* is then defined as
$$
\mbox{SQB} := \sum_{i=1}^{p} n_{i}\left(\bar{y}_{i}-\bar{y}\right)^{2},
$$ {#eq-sqb}
where 
$$
\bar{y} := \frac{1}{n} \sum_{i=1}^{p} \sum_{j=1}^{n_{i}} y_{ij} 
$$ {#eq-bar-y}
denotes the grand mean (mean across all data points in all groups) and  
$$
\bar{y}_{i}=\frac{1}{n_{i}} \sum_{j=1}^{n_{i}} y_{ij}
$$ {#eq-bar-y-i}
denotes the mean of the data points in the $i$th group. Large values of SQB 
imply a strong dependence of the group sample means on the respective treatment level 
$i$, while small values of SQB imply a weak dependence on the respective treatment level.
In this sense, SQB represents the data variability explained by considering the 
respective treatment level. Note that $\mbox{SQB}$ formula weighs the squared 
deviation between the $i$th group mean and the grand mean by the number of 
data points $n_i$ in the $i$th group. Deviations in larger groups thus contribute 
more to the $\mbox{SQB}$ than deviations in smaller groups. In *balanced designs*,
for which by definition all group sizes are identical, all groups contribute equally. 

The *within-sum-of-squares* is defined as
$$
\mbox{SQW} := \sum_{i=1}^{p} \sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}.
$$
Here, the sum of squared deviations of all data points from their group-specific
mean within a given group $\sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}$
is summed across all $p$ groups. SQW thus represents the data variability summed 
across all treatment levels that remains after the variability in the $i$th group 
has been explained by its respective group mean. Note that in the $\mbox{SQB}$ the
factor $n_i$ effectively scales the squared deviation between a group mean and 
the grand mean to account for the missing summation over $j = 1,...,n_i$ in $\mbox{SQW}$.

**Mean-sum-of-squares and $F$-value.** Dividing the $\mbox{SQB}$ and the $\mbox{SQW}$ 
by the number of contributing data points to obtain an average squared deviation 
per data point yields the so-called *mean between-sum-of-squares* and 
*mean-between-sum-of-squares*
$$
\mbox{MSB}:=\frac{\mbox{SQB}}{p-1} \mbox{ and }
\mbox{MSW}:=\frac{\mbox{SQW}}{n-p}.
$$ {#eq-msb-msw}
The denominators in @eq-msb-msw can be understood by realizing, that approximately 
$p$ "data points", i.e. group means contribute to the $\mbox{MSB}$ and approximately 
$n$ data points contribute to the $\mbox{SQW}$. The exact numbers in the denonminators
are referred to as *degrees of freedom*, such that the *between-group degrees of freedom (DFB)*
are given by $p-1$ and the *within-group degrees of freedom (DFW)* are given by $n-p$.

The *$F$-value* is defined as the ratio of $\mbox{MSB}$ and $\mbox{MSW}$, 
$$
F = \frac{\mbox{MSB}}{\mbox{MSW}}.
$$ {#eq-F-value}
$F$-values larger than $1$ thus indicate that the $\mbox{MSB}$ is larger than the 
$\mbox{MSW}$. In this sense, $F$-values larger than $1$ thus indicate that the 
averaged treatment-induced group mean variability is larger than the data point 
variability at a constant treatment level. The following **R** code evaluates the 
necessary quantities for computing the $F$-value for the example data sets 
of @tbl-anova-data-set-1 and @fig-anova-data-set-1.

\tiny
```{r, echo = T}
D       = read.csv("./_data/anova-data-set-1.csv")                              # dataframe in long format                                                
Y       = as.matrix(data.frame(split(D$Collagen, D$Treatment)))                 # wide format as simple matrix  
y_bar   = mean(Y)                                                               # grand mean 
y_bar_i = colMeans(Y)                                                           # group means
DFB     = ncol(Y) - 1                                                           # p -1
DFW     = length(Y) - ncol(Y)                                                   # n - p
SQB_i   = (y_bar_i - y_bar)^2                                                   # SQB terms
SQB     = sum(SQB_i * nrow(Y))                                                  # SQB 
SQW_i   = colSums((Y-matrix(y_bar_i,nrow(Y),length(y_bar_i), byrow = TRUE))^2)  # SQW terms
SQW     = sum(SQW_i)                                                            # SQW
MSB     = SQB/DFB                                                               # MSB  
MSW     = SQW/DFW                                                               # MSW
Eff     = MSB/MSW                                                               # F-value       
```
\normalsize
The results are as follows.
\tiny
```{r, echo = F}
# Ausgabe
cat("y_bar   :"  , round(y_bar,3),
    "\ny_bar_i :", round(y_bar_i,3),
    "\nDFB     :", round(DFB,3),
    "\nDFW     :", round(DFW,3),
    "\nSQB_i   :", round(SQB_i,3),
    "\nSQB     :", round(SQB,3),
    "\nSQW_i   :", round(SQW_i,3),
    "\nSQW     :", round(SQW,3),
    "\nMSB     :", round(MSB,3),
    "\nMSW     :", round(MSW,3),
    "\nF       :", round(Eff,3))
```

\normalsize
Obviously, the group means deviate significantly from the grand mean, while the
deviation of each data point from its respective group mean is not too large. Together,
this results in an $F$-value that is a lot larger than 1. 

For the example data set of @tbl-anova-data-set-2 and @fig-anova-data-set-2, 
the results are as follows.

\tiny
```{r, echo = F}
D       = read.csv("./_data/anova-data-set-2.csv")                              # dataframe in long format                                                
Y       = as.matrix(data.frame(split(D$Collagen, D$Treatment)))                 # wide format as simple matrix  
y_bar   = mean(Y)                                                               # grand mean 
y_bar_i = colMeans(Y)                                                           # group means
DFB     = ncol(Y) - 1                                                           # p -1
DFW     = length(Y) - ncol(Y)                                                   # n - p
SQB_i   = (y_bar_i - y_bar)^2                                                   # SQB terms
SQB     = sum(SQB_i * nrow(Y))                                                  # SQB 
SQW_i   = colSums((Y-matrix(y_bar_i,nrow(Y),length(y_bar_i), byrow = TRUE))^2)  # SQW terms
SQW     = sum(SQW_i)                                                            # SQW
MSB     = SQB/DFB                                                               # MSB  
MSW     = SQW/DFW                                                               # MSW
Eff     = MSB/MSW                                                               # F-value       
```
```{r, echo = F}
# Ausgabe
cat("y_bar   :"  , round(y_bar,3),
    "\ny_bar_i :", round(y_bar_i,3),
    "\nDFB     :", round(DFB,3),
    "\nDFW     :", round(DFW,3),
    "\nSQB_i   :", round(SQB_i,3),
    "\nSQB     :", round(SQB,3),
    "\nSQW_i   :", round(SQW_i,3),
    "\nSQW     :", round(SQW,3),
    "\nMSB     :", round(MSB,3),
    "\nMSW     :", round(MSW,3),
    "\nF       :", round(Eff,3))
```

\normalsize
For this data set, the individual group means do not vary much about the grand
mean, such that the $\mbox{SQB}$ is not large in comparison to the $\mbox{SQW}$
and the $F$-value thus smaller than 1. 

In **R**, the in-built function `aov()` can be used to obtain the same results.
In the following code, the `aov(D$Collagen ~ D$Treatment, data = D)` syntax means that the data
in the dataframe field `Collagen` should be grouped according to the information 
in the dataframe field `Treatment`. The generic `summary()` function than prints
results of the computation to the command window. Here `D$Treament` corresponds 
to the *between* quantities and `Residuals` corresponds to the *within* quantities.
The meaning of `Pr(>F)` and the associated Significance codes will become clearer after
in @sec-anova-as-a-linear-model.

\tiny
```{r, echo = T}
D       = read.csv("./_data/anova-data-set-1.csv")                              # data set 1 
res.aov = aov(D$Collagen ~ D$Treatment, data = D)                               # ANOVA as a computational procedure
summary(res.aov)                                                                # results 
```

```{r, echo = T}
D       = read.csv("./_data/anova-data-set-2.csv")                              # data set 2
res.aov = aov(D$Collagen ~ D$Treatment, data = D)                               # ANOVA as a computational procedure
summary(res.aov)                                                                # results  
```
\normalsize

**Partitioning of the Total-Sum-Squares**. The $\mbox{SQB}$ and $\mbox{SQW}$
are mutually dependent quantities, in the sense that their sum corresponds to the 
*total-sum-squares* of a given data set. The total-sum-of-squares corresponds to 
the squared deviations of all data points regardless of group membership from 
the grand mean,   
$$
\mbox{SQT} := \sum_{i=1}^{p} \sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}\right)^{2}.
$$ {#eq-sqt}
As shown in @sec-total-sum-of-squares-partition, it holds that
$$
\mbox{SQT} = \mbox{SQB}  + \mbox{SQW}.
$$ {#eq-sqt-partition}
If one understands the $\mbox{SQB}$ as a measure of the treatment-induced variability
and the $\mbox{SQW}$ as a measure of the data set's inherent unsystematic and random 
variability, @eq-sqt-partition can intuitively be read as 
$$
\mbox{Total data variability} = \mbox{Treatment-induced data variability}  + \mbox{Unsystematic data variability}.
$$ {#eq-sqt-partition-intuition}
@eq-sqt-partition-intuition then forms the basis for understanding ANOVA from a linear
model perspective as discussed next.

# ANOVA as a linear model {#sec-anova-as-a-linear-model}
## Model formulation {#sec-model-formulation}

A deeper understanding of the ANOVA approach is afforded by viewing it as an 
instance of a *linear modelling approach* to data analysis. In contrast to the 
computational viewpoint described above, modelling approaches do not start with 
a data set at hand and merely constitute a set of formulas to summarize such data set, 
but first and foremost provide hypotheses about the generative processes that 
give rise to observed data. As such, modelling approaches subserve the theoretical 
and deductive side of science, as well as its empirical and inductive aspects. 
Linear models can be shown to underlie a wide variety of methods from classical statistics, 
such as $z$- and $t$-tests, one-factorial and multifactorial ANOVA, simple linear regression, multiple 
linear regression and the analysis of covariance (ANCOVA). They also form the 
backbone for more generalized  approaches to data analysis, such as multivariate, 
latent variable or nonlinear data modelling. A modern comprehensive introduction 
to linear models cannot eshew a detour through probability theory and some basics 
of linear algebra. However, for the purposes of this tutorial, attention is 
restricted to the structural features of the linear model in the context of ANOVA. 
The probabilistic components are addressed primarily through intuitive explanations, 
and references to matrix-based computations are deliberately omitted.

In essence, the linear modelling approach to ANOVA conceives each observed
data point as the sum of a *deterministic component* induced by the treatment 
at each level and an unsystematic *random component* induced by additional 
processes such as the particularities of the observational unit under study or
inadvertent variations in the experimental application of the respective 
treatment level:
$$
\mbox{Data} = 
\mbox{Deterministic treatment-induced component} + 
\mbox{Random unsystematic component}
$$ {#eq-linear-model-intuition}
Note the similarity of this idea to the decomposition of the total data variability
in @eq-sqt-partition-intuition. More formally, the basic model for $j$th data 
point in the $i$th treatment group of an ANOVA design takes the form
$$
y_{ij} = \mu_i + \varepsilon_{ij}.
$$ {#eq-y-ij}
Here, $\mu_i$ is a fixed, deterministic unknown number that depends on the treatment
level $i = 1,...,p$. $\varepsilon_{ij}$ is a random variable that is specific to the 
$j$th data point in the $i$th treatment group, hence the double index. $\varepsilon_{ij}$
is variably referred to as an *error variable*, a *measurement error*, or a *random fluctuation*. 
As a random variable, $\varepsilon_{ij}$ takes on values at random with prespecified 
probabilities. A standard assumption for the error terms in linear models is that 
they follow a normal distribution (cf. @sec-the-normal-distribution), that is, 
a Gaussian bell-shaped curve centered at zero with variance parameter $\sigma^2$. 
This means that if one generates multiple realizations of a random variable like 
$\varepsilon_{ij}$, these realizations have a higher probability to be located 
close to zero than being located far away from zero. Because under this assumption 
most of the time $\varepsilon_{ij}$ in @eq-y-ij takes on values close to zero, 
$y_{ij}$ most of the time takes on values close to $\mu_i$, rendering $\mu_i$ 
its so-called *expected value*. We will go a bit deeper into the significance 
of the $\varepsilon_{ij}$ component for the applied ANOVA procedure in @sec-inference. 
For the moment we focus on the deterministic component $\mu_i$ that we have identified 
as the expected value of the data point $y_{ij}$. 

While in principle using a single number $\mu_i$ for each treatment-level to reflect 
the deterministic data component of a data point $y_{ij}$ is perfectly fine, most
statistical software solutions use a different representation that has better 
generalisation properties to more complex data acquisition scenarios and fits the 
matrix computations underlying the evaluation of linear models in software better. This 
parameterization is commonly referred to as *reference group coding*. The idea 
of this approach is to encode the expected value of the data points in a 
specific group, referred to as the *reference group*, by a value $\mu_0$ and to 
only encode the *differences* in expected values between the treatment groups and
the reference group as additional parameters. Which group is chosen as the reference 
group is not of primary importance, when using statistical software, its usually the first 
group in a data table. However, if there exists a baseline group in which e.g.,
no treatment or a sham-treatment is applied, it is sensible to use this as the
reference group for the remaining parameters to be readily interpretable. In the 
reference group coding approach to ANOVA, the linear model for the data points
$y_{ij}$ takes the form
$$
y_{1j} = \mu_0 + \varepsilon_{1j}
\mbox{ and }
y_{ij} = \mu_0 + \alpha_i + \varepsilon_{ij}
\mbox{ for }
i = 2,...,p.  
$$ {#eq-anova-lm}
The values $\alpha_2,...,\alpha_p$ encode the differences in expected values between the $i$th 
group and the reference group. A fundamental aim of ANOVA as an inference approach 
is to infer whether the true values of $\alpha_2,...,\alpha_p$ are different 
from zero or not.  

## Parameter estimation {#sec-parameter-estimation}

@eq-anova-lm constitutes a model for a data set acquired in an ANOVA design.
As most models of quantitative phenomena, @eq-anova-lm has parameters that are 
*estimated* based on the observed data $y_{ij}$. To understand this, it is sensible
to reflect again on the data generative process that is encoded in @eq-anova-lm.
The idea is that the data values $y_{1j}, j = 1,...,n_1$ in the reference group 
are generated based on a fixed unknown value for the measured property in the
reference group, which is termed $\mu_0$. However, this value of $\mu_0$ is not
directly observed, but only after addition of the observational unit-specific random
term $\varepsilon_{1j}$. Likewise, the values of $\mu_0$ and $\alpha_i$ 
for the remaining groups are only observed after addition of their respective random
contributions $\varepsilon_{ij}$. In this setting, the values of $\mu_0$ and 
$\alpha_i$ are referred to as *true, but unknown, values*. In fact, even after an 
ANOVA is computed, the actual values of $\mu_0$ and $\alpha_i$ will remain unknown. 
However, during the execution of an ANOVA procedure *reasonable guesses* or 
*estimates* for $\mu_0$ and the $\alpha_i$'s are evaluated. 

The ANOVA parameter estimates can be motivated by noting that the random contributions 
are symmetrically distributed around zero under the assumption of their normality. 
If we consider for example the reference group with $n_1 = 5$ and assume that 
$\mu_0 = 5$ and $\varepsilon_{11} = 0$, $\varepsilon_{12} = 1$, $\varepsilon_{13} = -1$, 
$\varepsilon_{14} = 2$ $\varepsilon_{15} = -1$, we obtain the data points

\begin{center}
\begin{tabular}{l}
$y_{11} = \mu_0 + \varepsilon_{11} = 5 + 0 = 5$ \\
$y_{12} = \mu_0 + \varepsilon_{12} = 5 + 1 = 6$ \\
$y_{13} = \mu_0 + \varepsilon_{13} = 5 - 1 = 4$ \\
$y_{14} = \mu_0 + \varepsilon_{14} = 5 + 2 = 7$ \\
$y_{15} = \mu_0 + \varepsilon_{15} = 5 - 1 = 4$ \\
\end{tabular}
\end{center}

As positive and negative contributions to $\mu_0$ cancel each other out when 
computing an average across these $y_{1i}$'s, a reasonable procedure to estimate
$\mu_0$ is to use the mean of the $y_{1i}$ as an estimate for $\mu_0$.
Estimates for true, but unknown, parameters are usually denoted with a little hat.
Thus, the usual estimate of $\mu_0$ in this scenario is 
$$
\hat{\mu}_0 = \frac{1}{n_1}\sum_{j=1}^{n_1} y_{1j},
$$
which for the data shown above evaluates to $\hat{\mu}_0 = 5.2$. This value is 
obviously not identical with the true, but unknown value of $\mu_0$, but also not
too far off. The statistical theory of linear models indeed assures that in the one-way
ANOVA scenario with reference group coding, the formation of group averages is 
a reasonable approach to estimate the parameters $\mu_0$, $\alpha_2$, ..., $\alpha_p$. 
Specifically, it can be shown that a reasonable estimator for $\mu_0$ is indeed 
the mean of the reference group and the differences between the respective group 
means and the mean of the reference group are reasonable estimators for 
$\alpha_2$, ..., $\alpha_p$, formally,
$$
\hat{\mu}_0 = \bar{y}_1 = \frac{1}{n_1}\sum_{j=1}^{n_1} y_{1j} 
\mbox{ and }
\hat{\alpha}_i 
= \bar{y}_i - \bar{y}_1 
= \frac{1}{m}\sum_{j=1}^{m} y_{ij} - \frac{1}{n_1}\sum_{j=1}^{n_1} y_{1j}
\mbox{ for } i = 1,...p.
$$ {#eq-parameter-estimation}

Note that the estimation of $\alpha_i$ as the difference between the $i$th group mean
and the reference group mean is well in line with the interpretation of $\alpha_i$
as encoding the difference in expected value between the $i$th group values and the 
reference group values.

In **R** parameter estimates such as $\hat{\mu}_0$ and $\hat{\alpha}_2$,..., $\hat{\alpha}_p$ 
are routinely evaluated when using the linear modelling approach
to ANOVA. The generic function to apply all sorts of linear models to data in **R** 
is `lm()`. To execute an ANOVA as described above, the syntax `lm(D$Collagen ~ D$Treatment, data = D)`
here again means that the data in the dataframe field `Collagen` should be grouped
according to the information in the dataframe field `Treatment`. When calling `lm()`
in this manner, the respective linear model is formulated and, based on the data
at hand, the parameter estimates are evaluated. These can be inspected in the 
`coefficients` field of the resulting **R** object. We demonstrate this for 
the data set in @tbl-anova-data-set-1. Note that the `(Intercept)` term $\hat{\mu}_0$ 
corresponds to  the mean of the first group (`T1`), which is used as reference group, 
and the `D$TreatmentT2`, `D$TreatmentT3`, `D$TreatmentT4` terms for $\hat{\alpha}_2$, 
$\hat{\alpha}_3$, and $\hat{\alpha}_4$ correspond to the differences of the 
respective group means and the reference group mean.

\tiny
```{r, echo = T}
# ANOVA as a linear model analysis
D         = read.csv("./_data/anova-data-set-1.csv")                            # data set 1
anova     = lm(D$Collagen ~ D$Treatment, data = D)                              # ANOVA as a computational procedure
print(anova$coefficients)                                                       # parameter estimates

# comparison with group means
Y         = as.matrix(data.frame(split(D$Collagen, D$Treatment)))               # wide format as simple matrix  
y_bar_i   = colMeans(Y)                                                         # group means 
mean_ref  = y_bar_i[1]                                                          # reference group mean
mean_ref                                                                        # print
mean_diff = y_bar_i[2:4] - y_bar_i[1]                                           # group mean differences
mean_diff                                                                        # print
```

\normalsize
## Inference {#sec-inference}

To understand the logic of *$p$-values* and related concepts such as *Frequentist
estimator properties*, *confidence intervals* and *error control in hypothesis testing*, 
it is crucial to remind oneself of the *Frequentist probability interpretation*  used 
standard applications of linear models. We have already noted that in the ANOVA 
model @eq-anova-lm the parameters $\mu_0$ and $\alpha_2,...,\alpha_p$ are conceived
as fixed unknown values that govern the data generating process of the ANOVA
scenario. For $i = 1,...,p$ and $j = 1,...,m$ the $\varepsilon_{ij}$'s, on the other
hand, are conceived as random variables. Random variables take on values with 
prespecified probabilities. For example, we can talk about the probability that 
the random variable $\varepsilon_{23}$ takes on a value in an interval such as 
$[-0.10,0.10]$. The probability for the error variable $\varepsilon_{23}$ 
to take on specific values also determines the probability of $y_{23}$ to take on 
specific values, given fixed values of $\mu_0$ and $\alpha_2$. For example, if 
$\mu_0 = 1$ and $\alpha_2 = 2$ and $\varepsilon_{23}$ has a high probability to take
on values in the interval $[-0.10,0.10]$, then $y_{23}$ has a high probability 
to take on values in the interval $[1 + 2 -0.10, 1 + 2 + 0.10] = [2.90,3.10]$. The
important thing here is to realize that given the fixed nature of the parameters
$\mu_0$ and $\alpha_i, i = 2,...,p$ and the random nature of the $\varepsilon_{ij}$'s
also the data variables are conceived as random variables.  

Now the fundamental assumption of the Frequentist approach to statistics is to assume
that a given data set, such as the data presented in @tbl-anova-data-set-1, reflects a single 
realization of each of the $\varepsilon_{ij}$'s and in turn the $y_{ij}$'s for 
true, but unknown, fixed parameters $\mu_0$ and $\alpha_i, i = 2,...,p$. From this
viewpoint, the collection of the $y_{ij}$'s in @tbl-anova-data-set-1 thus represents a single
joint realization of the random variables constituting the respective ANOVA
design. Against the background of the assumed random nature of the $\varepsilon_{ij}$'s 
involved, however, it is easily conceivable to imagine a second joint realization
of the random variables constituting the ANOVA design. In basic introductions
to statistics this is sometimes presented as "repeating the same study under identical
circumstances and collecting data that, by an large, up to random fluctation, resembles
that of the initial study". Yet, it is crucial to realize that the Frequentist assumption
underlying classical inference in linear models is just this, an assumption. In essence
it constitutes a cognitive model that is superimposed onto observed data and is 
no more "real" or "unreal" than any other form of human imagination. Nevertheless
superimposing the Frequentist assumption on a given experimental and data-analytical 
scenario allows for asking how observed data, or numbers computed from these data, 
such as means, sum-of-squares, mean sum-of-squares or $F$-values are distributed,
assuming known-properties of the random variables involved in the data generating
process, given here by the $\varepsilon_{ij}$'s. 

A straighforward way to gain intuition about the Frequentist distributional 
properties of data-analytical quantities is to simulate the repeated execution 
of a given study with identical true, but unknown, parameters on a computer. 
As this approach makes abundant use of random number generators, it is sometimes 
referred to as a *Monte Carlo approach* in reference to the famous casino in 
Monte Carlo. We prefer to refer to it as a *sampling approach*, as random 
variables are repeatedly sampled. For example, the following **R** code generates 
10 samples from a single normal random variable with expectation parameter $2$ 
and variance parameter $1$.

\tiny
```{r, echo = T}
n = 1                                                                           # number of random variables
s = 10                                                                          # number of samples taken
m = 2                                                                           # expectation parameter
v = 1                                                                           # variance parameter
for(i in 1:s){                                                                  # repeated sampling          
  print(rnorm(n,m,v))}                                                          # sampling a normally distributed random variable      
```
\normalsize 

To understand the notion of an *$F$-distribution* and the associated *$p$-value*
of an observed *$F$-value*, it is crucial to realize that in essence, an $F$-value 
is a fairly condensed representation of a data set as presented in @tbl-anova-data-set-1.
Everytime a new data set is realized (in simulation or reality), new grand
means, new group means, new sum-of-squares, new mean-sum-of-squares and a new 
$F$-value are realized. The fundamental logic of *$p$-values* is then to ask, how
probable a random $F$-value at least as extreme as an actual observed $F$-value under the 
assumption that, in the ANOVA design, all true, but unknown, group differences
with respect to the reference group $\alpha_2,...,\alpha_p$ are zero. This latter
assumption 
$$
\alpha_2 = \cdots = \alpha_p = 0
$$
is referred to as the *(global) null hypothesis* in an ANOVA design, as 
it represents the assumption of no treatment-induced effects. Note that the null
hypothesis is formulated with respect to the true, but unknown, values not their
empirical counterparts, such as the $\hat{\alpha}_2$,..., $\hat{\alpha}_p$. 
The logic of  Frequentist inference is then to conclude that if the probability 
$p$ of observing a random $F$-value at least as extreme, i.e., equal to or larger than,
the actual observed $F$-value is small, the null hypothesis should be rejected. 
This is the essence of the highly controversial, but very popular 
*null-hypothesis significance testing* approach of classical Frequentist 
statistics. In the following, we trace how the distribution of the $F$-value is 
constructed based on the normal assumption about the error variables in the ANOVA
linear model to illustrate the impact of the probabilistic assumptions in 
@eq-anova-lm for the practice of reporting $p$-values.

### Construction of the $F$-distribution {-}

The Frequentist distribution of the $F$-value under the null hypothesis of the 
ANOVA model derives from sequential transformations of the error variables 
$\varepsilon_{ij}$ for all $i = 1,...,p$ and $j = 1,...,n_i$. As discussed above, 
the error variables are assumed to be *identically* distributed according to a 
normal distribution with expectation parameter $0$ and variance parameter $\sigma^2$. 
Furthermore, the error variables are assumed to be *independently* distributed. 
In the context of normal distributions, this also implies that they are uncorrelated. 
Formally, this means that the distribution of a single $\varepsilon_{ij}$ can 
be described by the probability density function of the univariate normal distribution,
$$
p\left(\varepsilon_{ij}\right) := \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{1}{2\sigma^2}\left(\varepsilon_{ij}-\mu\right)^2\right)
$$
for $\mu:=0$. Intuitively, this means, that if a certain observational unit realizes
by chance for example a high value for its error variable, the probability for 
none of the other observational units to also realize a very high (or, 
for negative covariations, a very low) error variable value is affected in any way. 
The essence of the Frequentist distribution framework for the ANOVA model 
leading up to the $F$-value distribution is then to mathematically derive the 
formulas for the distributions of the transformed independent and identically normally 
distributed error variables through the various stages of the ANOVA model and 
the evaluation of the various statistics cumulating in the $F$-distribution. 
Tracing this process mathematically is beyond the scope of the current tutorial. 
We nevertheless list the core steps below and provide a sampling-based validation 
in @fig-anova-F-distribution. 

(1) The assumption of identically and independently normally distributed error 
variables, exemplarily shown for $\varepsilon_{11}$ in @fig-anova-F-distribution\unskip A, 
results in the fact that under the ANOVA model, the data variables $y_{ij}$ 
are independently normally distributed with the same variance as the error variables,
examplarily shown in @fig-anova-F-distribution\unskip B, as adding constant numbers to 
normally distributed random variables results in another normally distributed 
random variable.

(2) The normal distribution of the data variables $y_{ij}$ in turn implies the 
normal distribution of the group means, exemplarily shown for $\bar{y}_1$ in
@fig-anova-F-distribution\unskip C  and the grand mean, as shown in @fig-anova-F-distribution\unskip D, 
as summing together normally distributed random variables and multiplying them 
by a constant renders the result normally distributed.

(3) Subtracting normally distributed random variables from each other yields yet 
other normally distributed random variables, such that the deviations in the 
$\mbox{SQB}$ and the $\mbox{SQW}$ are themselves normally distributed, as shown 
exemplarily for $y_{11} - \bar{y}_1$ and $\bar{y}_1 - \bar{y}$ in @fig-anova-F-distribution\unskip E
and @fig-anova-F-distribution\unskip F, respectively.

(4) Squaring normally distributed random variables, in the current case the 
deviations between individual data variables and their group mean or the group 
means and the grand mean, does not result in normally distributed random variables. 
This is readily appreciated by considering a random variable centered around zero 
and considering the square of its realizations - all negative realizations will
become positive, values between 0 and 1 will move closer to zero and values 
above 1 will move further away from 1. Squaring thus breaks the symmetry of 
normally distributed values. Broadly speaking, the square of an independent 
normally distributed random variable, as well as the sum of such squares, 
follows a $\chi^2$ distribution. Unlike the normal distribution, the corresponding 
probability density function has a different — yet analytically available — form.
In the context of the $F$-distribution, the variance-scaled versions of $\mbox{SQB}$ 
and $\mbox{SQW}$ are independently $\chi^2$-distributed with parameters $p-1$ and $n-p$,
respectively, as shown in @fig-anova-F-distribution\unskip G and @fig-anova-F-distribution\unskip H.

(5) Finally, forming the ratio of two independent $\chi^2$-distributed random 
variables divided by their respective parameters leads to a distribution with 
probability density function referred to as the $F$-distribution, as shown in 
@fig-anova-F-distribution\unskip I.

```{r, echo = F, eval = F}
library(MASS)                                                                   # multivariate normal distribution
set.seed(2)                                                                     # reproducibility 
m       = 10                                                                    # number of observational units per group
p       = 4                                                                     # number of groups 
n       = p*m                                                                   # number of observation units 
Xt        = cbind(                                                              # design matrix
             matrix(1,nrow = n, ncol = 1),                                      # design matrix
             kronecker(diag(p), matrix(1,nrow = m,ncol = 1)))                   # design matrix
X       = Xt[,-2]                                                               # design matrix
I_n     = diag(n)                                                               # n x n identity matrix
beta    = matrix(c(5,0,0,0), nrow = p)                                          # \beta in \Omega_0
sigsqr  = 4                                                                     # \sigma^2   

# sample arrays
nsamp   = 100000                                                                # number of samples
eps_11  = rep(NaN, nsamp)                                                       # error variable
y_11    = rep(NaN, nsamp)                                                       # data variable
y_bar   = rep(NaN, nsamp)                                                       # grand mean   
y_bar_1 = rep(NaN, nsamp)                                                       # group mean
SQB     = rep(NaN, nsamp)                                                       # between sum of squares
SQW     = rep(NaN, nsamp)                                                       # within  sum of squares
Eff     = rep(NaN, nsamp)                                                       # F-values
for(s in 1:nsamp){                                                              # sampling iterations
  eps         = mvrnorm(1, rep(0,n),sigsqr*I_n)                                 # error vector
  eps_11[s]   = eps[1]                                                          # error variable
  y           = X %*% beta + eps                                                # data vector    
  y_11[s]     = y[1]                                                            # data variable
  D           = data.frame(TR = rep(paste0("T", 1:4), each = m), CG = y)        # dataframe
  Y           = as.matrix(data.frame(split(D$CG, D$TR)))                        # wide format as simple matrix  
  y_bar[s]    = mean(Y)                                                         # grand mean 
  y_bar_i     = colMeans(Y)                                                     # group means
  y_bar_1[s]  = y_bar_i[1]                                                      # group means
  DFB         = ncol(Y) - 1                                                     # p -1
  DFW         = length(Y) - ncol(Y)                                             # n - p
  SQB_i       = (y_bar_i - y_bar[s])^2                                          # SQB terms
  SQB[s]      = sum(SQB_i * nrow(Y))                                            # SQB 
  SQW_i       = colSums((Y-matrix(y_bar_i,nrow(Y),length(y_bar_i),byrow=TRUE))^2) # SQW terms
  SQW[s]      = sum(SQW_i)                                                      # SQW
  MSB         = SQB[s]/DFB                                                      # MSB  
  MSW         = SQW[s]/DFW                                                      # MSW
  Eff[s]      = MSB/MSW                                                         # F-value
}
```

```{r, eval = F}
# visualization
library(latex2exp)
pdf(                                                                             
file        = "./_figures/anova-$F$-distribution.pdf",                
width       = 14,                                                                     
height      = 14)                                
par(                                                                    
family      = "sans",                                                  
mfcol       = c(3,3),
mar         = c(4,4,2,1), 
oma         = c(1,1,3,1),                                                  
pty         = "m",                                                     
bty         = "l",                                                     
lwd         = 1,                                                       
las         = 1,                                                       
mgp         = c(2,1,0),                                                
xaxs        = "i",                                                   
yaxs        = "i",                                                  
font.main   = 1,                                                       
cex         = 1.2,
cex.main    = 1.2)

# analytical distributions
samples = 
list(
eps_11, 
y_11, 
y_bar_1, 
y_bar, 
y_11 - y_bar_1, 
y_bar_1 - y_bar,
SQB/sigsqr,
SQW/sigsqr,
Eff)
values = list(
seq(-5,5,len = 1000),
seq(0,10,len = 1000),
seq(0,10,len = 1000),
seq(0,10,len = 1000),
seq(-5,5,len = 1000),
seq(-8,8,len = 1000),
seq(0,10,len = 1000),
seq(0,100,len = 1000),
seq(0,10,len = 1000))
pdfs = 
list(
dnorm(values[[1]], 0, sqrt(sigsqr)),
dnorm(values[[2]], beta[1], sqrt(sigsqr)),
dnorm(values[[3]], beta[1], sqrt(sigsqr/m)),
dnorm(values[[4]], beta[1], sqrt(sigsqr/n)),
dnorm(values[[5]], 0, sqrt(sigsqr*(1 - 1/m))),
dnorm(values[[6]], 0, sqrt(sigsqr/m - sigsqr/n)),
dchisq(values[[7]], p-1),
dchisq(values[[8]], n-p),
df(values[[9]], p-1, n-p))
tlabs = 
c(
TeX('$\\epsilon_{11}$'),
TeX('$y_{11}$'),
TeX('$\\bar{y}_{1}$'),
TeX('$\\bar{y}$'),
TeX('$y_{11} - \\bar{y}_1$'),
TeX('$\\bar{y}_{1} - \\bar{y}$'),
TeX('$SQB/\\sigma^2$'),
TeX('$SQW/\\sigma^2$'),
TeX('$F$'))
yulim   = c(0.25,0.25,0.7,1.4,0.25,0.8,0.30,0.06,0.8)
for(i in 1:9){
  hist(
  samples[[i]],                                                                       
  breaks      = 100,                                                                
  col         = "gray90",                                                           
  prob        = TRUE,                                 
  xlab        = "",                                                                
  ylab        = "",  
  ylim        = c(0, yulim[i]),                                                               
  main        = tlabs[i])
  lines(
  values[[i]],                                                                          
  pdfs[[i]],                                                                       
  lwd         = 2,                                                                  
  col         = "darkorange")  
}
dev.off()
```

![Construction of the $F$-distribution.](./_figures/anova-F-distribution-labeled.pdf){#fig-anova-F-distribution fig-align="center" width=90%}

### The $p$-value {-}

Because the $F$-value takes on larger values, if the true, but unknown, parameters 
$\alpha_2,...,\alpha_p$ take on values different from zero, asking how probable a 
given observed value of the $F$-value or a more extreme value is, gives a 
probability that becomes small for evidence against the null hypothesis. This 
probability is known as the *$p$-value* and can be visualized as the area under 
the curve of the $F$-distribution to the right of an observed $F$-value (@fig-anova-p-value). 
If this probability is low, for example lower than 0.05, the evidence in a data set 
is commonly interpreted as considerable evidence against the null hypothesis. 
Note however, that the $p$-value itself has no implications for the veracity or 
non-veracity of the null hypothesis: $F$-values at least as large as an observed 
$F$-value are just not very likely in the case that the null hypothesis is 
true, but they are by no means impossible. In fact, their associated probability 
is just the $p$-value. Also note that the existence of a $p$-value is crucially 
dependent on the normal distribution assumption of the error terms, as its value 
derives from the $F$-distribution, which is constructed based on exactly this 
distribution. In the **R** ANOVA table, the $p$-value is listed as `Pr(>F)`. With
the help of the cumulative distribution function of the $F$-distribution, which 
encodes the probability $\mathbb{P}(F \le f)$ for all observable $F$-values $f$, 
the $p$-value can be readily evaluated based on the following **R** code. @fig-anova-p-value
visualizes this process.

\tiny
```{r, echo = T}
p       = 4                                                                     # number of groups
m       = 10                                                                    # number of observational units per group
n       = p*m                                                                   # total number of data points    
F_value = 0.5                                                                   # observed $F$-value
p_value = 1-pf(F_value, p-1, n-p)                                               # associated p-value
cat("p-value: ", round(p_value,3))                                              # print
```
\normalsize

Finally, we note that the distribution of the $F$-value cannot only be evaluated under 
the null hypothesis $\alpha_2 = \cdots = \alpha_p = 0$, but equally well under any 
other hypothesis about the true, but unknown, parameters. This leads to the 
concept of the non-central $F$-distribution, an important tool in the context 
of ANOVA power analyses and optimal sample size calculations in the Frequentist setting.

```{r,echo = F, eval= F}
# PDF Speicherung
pdf(
file        =  "./_figures/anova-p-value.pdf",
width       = 9,
height      = 4.5)
par(
family      = "sans",
mfcol       = c(1,2),
pty         = "m",
bty         = "l",
lwd         = 1,
las         = 1,
mgp         = c(2,1,0),
xaxs        = "i",
yaxs        = "i",
font.main   = 1,
cex         = 1,
cex.main    = 1.2)

# Parameter
p           = 4                                                                 # Anzahl Gruppen
m           = 10                                                                # Anzahl experimenteller Einheiten pro Gruppe
n           = p*m                                                               # Gesamtstichprobengröße
F_value     = 0.5                                                               # F-value
eff         = seq(0,5,length=1e4 )                                              # F-Statistikwerte
Peff        = pf(eff,p-1,n-p)                                                   # F-Statistik KVF für H_0
peff        = df(eff,p-1,n-p)                                                   # F-Statistik WDF für H_0

# KVF Perspektive
plot(                                                                            # original density function
eff,
1 - Peff,
type        = "l",
xlab        = " ",
ylab        = " ",
ylim        = c(0,1),
main        = "Complementary CDF of F(3,36)") 
lines(
F_value,
0,
type        = "p",
pch         = 16,
xpd         = TRUE)
lines(
min(eff),
1 - p_value,
type        = "p",
pch         = 16,
xpd         = TRUE)
arrows(
x0          = F_value,
y0          = p_value,
x1          = min(eff),
y1          = p_value,
col         = "gray",
angle       = 35,
length      = .1)
arrows(
x0          = F_value,
y0          = 0,
x1          = F_value,
y1          = p_value,
col         = "gray",
angle       = 35,
length      = .1)
text(F_value, .1 , TeX("$F$-value"), xpd = TRUE)
text(-.6      ,p_value , TeX("p-value"), xpd = TRUE)

# WDF Perspektive
plot(
eff,
peff,
type        = "l",
ylab        = " ",
xlab        = " ",
ylim        = c(0,.8),
main        = "PDF of F(3,36)")
polygon(
c(eff[eff  >= F_value] , max(eff), F_value),
c(peff[eff >= F_value],       0, 0),
col = "gray90",
border = NA)
lines(
F_value,
0,
type        = "p",
pch         = 16,
xpd         = TRUE)
text(F_value,0.05, TeX("$F$-value") , xpd = TRUE)
text( 1.2, .15, TeX("p-value"), xpd = TRUE)
dev.off()
```

![Evaluation of the ANOVA $p$-value for the example data set 2 with observed $F$-value 0.5.](./_figures/anova-p-value.pdf){#fig-anova-p-value fig-align="center" width=90%}

\newpage
# ANOVA as a linear mixed model {#sec-anova-as-a-linear-mixed-model}

Thus far, we have assumed that the observational units aside from being assigned
to different treatment groups do not exhibit any predetermined inherent structure.
In fact, in @tbl-anova-data-set-1 and @tbl-anova-data-set-2, we have separated the 
inherent structure of observational units, i.e., fibroblasts from one cell culture
strain vs. fibroblasts from another cell culture strain, into two independent 
experiments and found evidence for treatment effects in one strain, but not the other. 
If there exist inherent structure in the observational units that is in one way or another
superimposed onto the structure of the experimental manipulation, it often pays to account for this 
structure in the modelling of the experimental data set. Today, such structure 
modelling is often achieved using *linear mixed models*, which in addition to 
(fixed) experimental treatment effects and (random) unsystematic effects also account 
for (random) observational unit effects that reflect some form of structure. In fact, 
not accounting for structured variation of experimental data that is induced by the observational units, but treating all
observational units as if they were independent, can have effects on the results
of ANOVA analyses. The reason for this is that the analytical form of the
$F$-distribution is built on the assumption of independent and identically distributed
error variables for all observational units. If this assumption is violated because 
certain error correlations exist, the actual distribution of $F$-values as computed 
by the procedures discussed in @sec-anova-as-a-computational-procedure may differ 
from the analytically available form as discussed in @sec-anova-as-a-linear-model. 
While in the past, these instances were usually dealt with by adapting the analytical 
form of the $F$-distribution appropriately, e.g., by evaluating modified degrees of 
freedom parameters for it as in the *Huyhn-Feldt* or *Greenhouse-Geisser corrections*, 
these methods have by and large been superseeded by linear mixed model approaches. 
To motivate and explore these approaches, we consider two scenarios of inherent observational 
unit structure that, if analysed using the conventional ANOVA approach assuming 
independent  observational unit error variables, lead to an increase of false negative results (loss of 
sensitivity) or an increase of false positive results. These scenarios are 
referred to as *randomized block designs* and *nested subsampling designs*. 

![Exemplary randomized block and nested subsampling designs. Both example designs
comprise four experimental treatments. In the randomized block design, the observational 
units within an experimental unit (block) are subjected to all treatments. 
Differences between experimental units thus cancel out, if differences
between the treatments are assessed. However, the spread of measurements in one 
condition may be larger than in a one-way ANOVA design with one experimental unit 
due to the varying experimental unit effects. In the nested subsampling design, observational
units of different experimental units are used across the different treatments. If, 
by chance, the observational units of experimental units E10 to E12 have high values due
to the nature of the experimental units E10 to E12, this may be mistaken as an 
effect of T4 if not accounted for.](./_figures/anova-randomized-block-nested-subsampling-designs.pdf){#fig-anova-randomized-block-nested-subsampling-designs fig-align="center" width=70%}

**Randomized block designs** and the closely related
*repeated-measures designs* are characterized by the fact that all experimental 
conditions are measured within a given experimental unit. In the cell biological example, 
if three different cell culture strains were used and each of the four experimental treatments 
was applied within each strain, each to a set of fibroblasts from the particular strain, 
this setup would be referred to as a randomized block design. Another common example are 
cognitive psychology experiments, in which behavioral responses of human participants 
are typically measured in response to all experimental treatments of interest. @fig-anova-randomized-block-nested-subsampling-designs\unskip A depicts the typical 
design of an ANOVA randomized block design. Randomized block designs have the benefit, 
that, at least in theory, data variation due to the experimental unit is equally 
expressed across all experimental treatments, such that observed differences between 
experimental treatments are not affected by the experimental units. They have the 
disadvantage that the effect of the experimental units may increase the data spread
about each treatment group mean.  

A basic linear mixed model for a randomized block design takes the form
$$
y_{kij} = \mu_0 + \alpha_i + \mu_k + \varepsilon_{kij},
$$ {#eq-anova-randomized-block}
where for $k = 1,...,n_k$, $i = 1,...,p$ and $j = 1,...,n_{ki}$ 

* $y_{kij}$ is the data of the $j$th observational unit in the $i$th treatment group in the $k$th experimental unit 
* $\mu_0$ is the average reference group effect across experimental units, 
* $\alpha_i$ is the average difference between treatment level $i$ and the average reference group effect with $\alpha_1 := 0$
* $\mu_k \sim N(0,\sigma_k^2)$ is the random effect of the $k$th experimental unit, 
* $\varepsilon_{kij} \sim N(0, \sigma_{\varepsilon}^2)$ is the error of the $j$th observational unit in the $i$th treatment in the $k$th experimental unit.

Note that in contrast to @eq-anova-lm, @eq-anova-randomized-block includes explicit
terms that reference the experimental unit, in particular the normally distributed
random experimental unit effect $\mu_k \sim N(0,\sigma_k^2)$. Parameter estimation 
for linear mixed models such as @eq-anova-randomized-block is complicated by the 
fact that in contrast to @eq-anova-lm, linear mixed models include multiple
latent random variables such as $\mu_k$ and $\varepsilon_{kij}$ and multiple parameters 
such as $\sigma_k^2$ and $\sigma_\varepsilon^2$. 

@fig-anova-randomized-block-design-data shows simulated data from a randomized 
block design with $k = 3$ experimental units `E1`, `E2`, `E3`, $p = 4$ treatments 
`T1`,`T2`,`T3` and `T4` and $n_{ki} = 10$ observational units per treatment group in each experimental unit. 
The data is generated based on the null hypothesis parameter setting $\alpha_2 = \alpha_3 = 0$.
Note that when averaging for example the data points of `T1` across all experimental units,
and assessing the remaining differences between the data points of `T1` across
all experimental units, this component of the within-sum-of-squares will be larger
than compared to a scenario in which only data from one of the experimental units 
is assessed. On the other hand, the observable differences across all treatment groups
across experimental units cancel out, if differences between treatment means are
evaluated in a per experimenal unit manner.

```{r, echo = F, eval = F}
library(MASS)                                                                   # multivariate normal distribution
library(nlme)                                                                   # linear mixed models
set.seed(0)                                                                     # reproducibility 
s           = 3                                                                 # number of cell culture strains
p           = 4                                                                 # number of groups per cell culture strain  
m           = 10                                                                # number of observational units per group per cell culture strain
k           = p*m                                                               # number of observation units per cell culture strain
n           = s*p*m                                                             # number of observational units in total
b           = s*p                                                               # number of groups in total 
X_t         = cbind(                                                            # fixed  effects design matrix construction
              matrix(1, nrow = k, ncol = 1),                                    # fixed  effects design matrix construction
              kronecker(diag(p), matrix(1,nrow = m,ncol = 1)))                  # fixed  effects design matrix construction
X_i         = X_t[,-2]                                                          # fixed  effects design matrix for one cell culture strain
X           = kronecker(matrix(rep(1,s), nrow = s), X_i)                        # fixed  effects design matrix
Z           = kronecker(diag(s), matrix(rep(1,k), nrow = k))                    # random effects design matrix
beta        = matrix(c(5,0,0,0), nrow = p)                                      # fixed effects null hypothesis parameter
mu_y        = X %*% beta                                                        # marginal expectation
sig_eps     = 3                                                                 # error variance
sig_b       = 3                                                                 # random effects variance
Sigma_y     = sig_b*(Z %*% t(Z)) + sig_eps*diag(n)                              # marginal covariance
y           = mvrnorm(1, mu_y, Sigma_y)                                         # data sampling
T           = rep(rep(paste0("T", 1:p), each = m), s)                           # treatment factor
E           = rep(paste0("E", 1:s), each = p*m)                                 # experimental unit factor           
D           = data.frame(TRM = T, EU = E, y = y)                                # data formatting 
write.csv(D, "./_data/anova-randomized-block-design-data-set.csv", row.names = FALSE) # 
```

```{r, eval = F}
# visualization as "stripchart"
library(latex2exp)
pdf(                                                                            
file        = "./_figures/anova-randomized-block-design-data.pdf",                                           
width       = 10,                                                                     
height      = 6) 
par(                                                                    
family      = "sans",                                                  
mfcol       = c(1,1),                                                  
pty         = "m",                                                     
bty         = "l",                                                     
lwd         = 1,                                                       
las         = 2,                                                       
mgp         = c(2,1,0),                                                
xaxs        = "i",                                                   
yaxs        = "i",                                                  
font.main   = 1,                                                       
cex         = 1.2,
cex.main    = 1.4,
xpd         = FALSE)
xpos        = 1:b
xjit        = rep(xpos, each = m) + runif(n, -0.15, 0.15)
tcols       = c("T1"="gray20","T2"="gray40","T3"="gray60","T4"="gray80")
ptcol       = adjustcolor(tcols[D$TRM], alpha.f = 0.9)
plot(
xjit, 
D$y, 
pch         = 19, 
col         = ptcol,
xaxt        = "n", 
xlab        = " ",
ylab        = "Observation", 
main        = "Randomized block design data",
xlim        = c(0.5,12.5),
ylim        = c(0,8))
axis(
1, 
at          = unique(xpos), 
labels      = rep(paste0("T", 1:p),s) ,
las         = 1)
abline(
v           = c(4.5, 8.5), 
lty         = 3, 
col         = "grey70")
mtext(
c("E1","E2","E3"), 
side        = 1, 
line        = 3,
at          = c(2.5,6.5,10.5), 
cex         = 1.4, 
las         = 1)
dev.off()
```

![Randomized block design data example.](./_figures/anova-randomized-block-design-data.pdf){#fig-anova-randomized-block-design-data fig-align="center" width=80%} 

Formally, it can be shown that the presence of random terms such as $\mu_k$ induces 
correlations in the marginal error and data covariance matrices. In the case of 
randomized block designs, these correlations induce an increase of the within-sum-of-squares 
such that, when not accounted for, on average lower $F$-values are observed than 
predicted based on independent observational unit error terms. 
@fig-anova-randomized-block-design-f-simulation\unskip B shows this shift of observed
$F$-values under the null hypothesis when computed based on samples from 
@eq-anova-randomized-block as opposed to computed based on samples from @eq-anova-lm as shown in @fig-anova-randomized-block-design-f-simulation\unskip A. 

The popularity of linear mixed models in the analysis of modern ANOVA designs
owes much to the availability of software packages that make data analysis with 
mathematically fairly complex models very easy in research applications. 
The most popular **R** packages for linear mixed model analysis are `nlme` (@pinheiro2000)
and `lme4` (@bates2004). We here demonstrate a linear mixed model analysis using
the `lme()` function of `nlme` which capitalizes on a *restricted maximum-likelihood*
approach for covariance component estimation. In addition to 
specifying which data should be modeled with respect to which experimental factor,
linear mixed model implementation typically required additional information
about the observational unit structure (grouping). For the data shown in 
@fig-anova-randomized-block-design-data, this information is encoded in the 
`EU` (experimental unit) column of the data set table. The following code snippet shows
the analysis of this data set using the linear mixed model of @eq-anova-randomized-block.

\tiny
```{r, echo = T, message = F, warning = F}
library(nlme)                                                                   # (non)linear mixed model package
D       = read.csv("./_data/anova-randomized-block-design-data-set.csv")        # dataframe in long 
anova(lme(y ~ TRM, data = D, random = ~ 1|EU))                                  # LMM treatment effect evaluation
```
\normalsize

Using this approach in the Frequentist simulations of @fig-anova-randomized-block-design-f-simulation\unskip C
shows that an analytical $F$-distribution is available that conforms to the observed
$F$-value distribution of the linear mixed model. Using a linear mixed model that 
properly accounts for the induced error correlations thus renders an ANOVA analysis 
of data from randomized block design more sensitive. 

```{r, echo = F, eval = F}
library(MASS)                                                                   # multivariate normal distribution
library(nlme)                                                                   # linear mixed models
set.seed(2)                                                                     # reproducibility 
s           = 3                                                                 # number of cell culture strains
p           = 4                                                                 # number of groups per cell culture straing 
m           = 10                                                                # number of observational units per group per cell culture strain
k           = p*m                                                               # number of observation units per cell culture strain
n           = s*p*m                                                             # number of observational units in total
X_t         = cbind(                                                            # fixed  effects design matrix construction
              matrix(1, nrow = k, ncol = 1),                                    # fixed  effects design matrix construction
              kronecker(diag(p), matrix(1,nrow = m,ncol = 1)))                  # fixed  effects design matrix construction
X_i         = X_t[,-2]                                                          # fixed  effects design matrix for one cell culture strain
X           = kronecker(matrix(rep(1,s), nrow = s), X_i)                        # fixed  effects design matrix
Z           = kronecker(diag(s), matrix(rep(1,k), nrow = k))                    # random effects design matrix
beta        = matrix(c(5,0,0,0), nrow = p)                                      # fixed effects null hypothesis parameter
sig_eps     = 3                                                                 # error variance
sig_b       = 3                                                                 # random effects variance
nsamp       = 10000                                                             # number of samples per model
I_n         = diag(n)                                                           # error covariance matrix structure      
F_lm        = rep(NaN, nsamp)                                                   # F-values for the linear model
F_lmm       = rep(NaN, nsamp)                                                   # F-values for linear mixed model
F_lme       = rep(NaN, nsamp)                                                   # F-values for the linear mixed model estimated using lme    

# linear model simulations
mu_y        = X %*% beta                                                        # marginal expectation
Sigma_y     = sig_eps*I_n                                                       # marginal covariance
for(i in 1:nsamp){                                                              # sampling iterations
  y         = mvrnorm(1, mu_y, Sigma_y)                                         # data sampling 
  T         = rep(rep(paste0("T", 1:p), each = m), s)                           # treatment factor
  E         = rep(paste0("E", 1:s), each = p*m)                                 # experimental unit factor           
  D         = data.frame(TRM = T, EU = E, y = y)                                # data formatting 
  R         = lm(y ~ TRM, data = D)                                             # ANOVA
  F_lm[i]   = summary(R)$fstatistic[1]                                          # F-value
}

# linear mixed model simulations of a balanced randomized block design
mu_y        = X %*% beta                                                        # marginal expectation
Sigma_y     = sig_b*(Z %*% t(Z)) + sig_eps*I_n                                  # marginal covariance
for(i in 1:nsamp){                                                              # sampling iterations
  y         = mvrnorm(1, mu_y, Sigma_y)                                         # data sampling
  T         = rep(rep(paste0("T", 1:p), each = m), s)                           # treatment factor
  E         = rep(paste0("E", 1:s), each = p*m)                                 # experimental unit factor           
  D         = data.frame(TRM = T, EU = E, y = y)                                # data formatting 
  R         = lm(y ~ TRM, data = D)                                             # ANOVA
  F_lmm[i]  = summary(R)$fstatistic[1]                                          # F-value, inppropriate linear model 
  F_lme[i]  = anova(lme(y ~ TRM, data = D, random = ~ 1|EU))["TRM", "F-value"]  # F-value, appropriate linear mixed model
}
```

```{r, eval = F}
library(latex2exp)
pdf(                                                                             
file        = "./_figures/anova-randomized-block-design-f-simulation.pdf",                
width       = 15,                                                                     
height      = 5)                                                                                                                                         
par(                                                                    
family      = "sans",                                                  
mfcol       = c(1,3),                                                  
pty         = "m",                                                     
bty         = "l",                                                     
lwd         = 1,                                                       
las         = 1,                                                       
mgp         = c(2,1,0),                                                
xaxs        = "i",                                                   
yaxs        = "i",                                                  
font.main   = 1,                                                       
cex         = 1.2,
cex.main    = 1)

# analytical f distribution
f_min       = 0                                                                  
f_max       = 4                                                                  
f_res       = 1e3                                                                
f           = seq(f_min,f_max, len = f_res)                                     
p_F         = df(f,p-1,n-p)                                                          

# linear model generated histogram and density
hist(
F_lm,                                                                       
breaks      = 100,                                                                
col         = "gray90",                                                           
prob        = TRUE,                                                               
xlim        = c(f_min, f_max),                                                    
ylim        = c(0,0.8),                                                            
xlab        = "F-value",                                                                
ylab        = "",                                                                 
main        = "Independent observational units \n Linear model analysis")
lines(
f,                                                                          
p_F,                                                                       
lwd         = 2,                                                                  
col         = "darkorange")      

# linear mixed model generated histogram and density
hist(
F_lmm,                                                                       
breaks      = 100,                                                                
col         = "gray90",                                                           
prob        = TRUE,                                                               
xlim        = c(f_min, f_max),                                                    
ylim        = c(0,1.2),                                                            
xlab        = "F-value",                                                                
ylab        = "",                                                                 
main        = "Randomized block design \n Linear model analysis")
lines(
f,                                                                          
p_F,                                                                       
lwd   = 2,                                                                  
col   = "darkorange")   

# linear mixed model generated histogram and density
f_min       = 0                                                                  
f_max       = 4                                                                  
f_res       = 1e3                                                                
f           = seq(f_min,f_max, len = f_res)                                     
p_F         = df(f,p-1,n-p-(s-1))  
hist(
F_lme,                                                                       
breaks      = 100,                                                                
col         = "gray90",                                                           
prob        = TRUE,                                                               
xlim        = c(f_min, f_max),                                                    
ylim        = c(0,0.8),                                                            
xlab        = "F-value",                                                                
ylab        = "",                                                                 
main        = "Randomized block design \n Linear mixed model analysis")
lines(
f,                                                                          
p_F,                                                                       
lwd   = 2,                                                                  
col   = "darkorange")  
dev.off()                                                       
```

![Loss of sensitivity in a randomized block design when analyzed with a linear model ANOVA,
but not a linear mixed model ANOVA.](./_figures/anova-randomized-block-design-f-simulation.pdf){#fig-anova-randomized-block-design-f-simulation fig-align="center" width=100%}  


**Nested subsampling designs** are experimental designs sometimes encountered in biological 
studies (cf. @yu2022). Here, the basic idea is that each experimental unit provides
a set of observational units and treatments are applied to sets of experimental units.
For example, suppose four different cytokine treatments are tested. For each treatment, 
fibroblasts are harvested from three different animals, and collagen production is then 
measured in these fibroblasts. Because the fibroblasts are not independent samples 
but instead are nested within animals (which are themselves grouped by treatment), 
this setup is considered a nested subsampling design. Intuitively, nested subsampling
designs have the disadvantage that they can easily conflate experimental unit and
treatment effects. In @fig-anova-randomized-block-nested-subsampling-designs\unskip B, assume that by
chance the fibroblasts of experimental units E10 to E12 produce high levels of
Collagen, irrespective of any treatment. If observed under the nested subsampling
design, this random experimental unit effect can readily be mistaken as an effect
of the fourth treatment level. Nested subsampling designs are thus prone to 
produce false positive effects, if their inherent observational unit structure is
not accounted for.

A basic linear mixed model for a nested subsampling design takes the following form.

$$
y_{ikj} = \mu_0 + \alpha_i + \mu_{ik} + \varepsilon_{ikj},
$$ {#eq-anova-nested-subsampling}
where for $k = 1,...,n_k$, $i = 1,...,p$ and $j = 1,...,n_{ik}$ 

* $y_{ikj}$ is the data of the $j$th observational unit of the $k$th experimental unit in the $i$th treatment group,
* $\mu_0$ is the average reference group effect across experimental units, 
* $\alpha_i$ is the average difference between treatment level $i$ and the average reference group effect with $\alpha_1 := 0$
* $\mu_{ik} \sim N(0,\sigma_k^2)$ is the random effect of the $k$th experimental unit in the $i$th treatment group, 
* $\varepsilon_{ikj} \sim N(0, \sigma_{\varepsilon}^2)$ is the error of the $j$th observational unit in the $k$th experimental unit in the $i$th treatment.

Note that in contrast to @eq-anova-randomized-block, the random effects $\mu_{ik}$
are experimental unit *and* treatment-specific. 


@fig-anova-nested-subsampling-design-data shows simulated data from a nested subsamplig
design with $n_k = 12$ experimental units `E1` to `E12`, $p = 4$ experimental treatments
and $n_{ik} = 10$ observational units per experimental unit. Note that the elevated 
observations in `T4` may well be induced by the experimental unit effects of `E10` ,`E11`,
and `E12`. In general it is easily possible, that experimental unit effects inflate
the mean between-sum-of-squares in a linear model ANOVA of nested subsampling design data.
In analogy to the the analysis of the randomized block design data, the following 
code snippet shows the analysis of this data set using the linear mixed model 
of @eq-anova-nested-subsampling.

\tiny
```{r, echo = T, message = F, warning = F}
library(nlme)                                                                   # (non)linear mixed model package
D       = read.csv("./_data/anova-nested-subsampling-design-data-set.csv")      # dataframe in long format
anova(lme(y ~ TRM, data = D, random = ~ 1|EU))                                  # LMM treatment effect evaluation
```
\normalsize

```{r, echo = F, eval = F}
# simulation of a single data set
library(MASS)                                                                   # multivariate normal distribution
set.seed(2)                                                                     # reproducibility 
p           = 4                                                                 # number of experimental groups 
m           = 30                                                                # number of observational units per group  
a           = 3                                                                 # number of animals per group 
b           = p*a                                                               # number of animals in total
k           = m/a                                                               # number of observation units per animal
n           = p*m                                                               # number of observational units in total
X_t         = cbind(                                                            # fixed  effects design matrix construction
              matrix(1, nrow = n, ncol = 1),                                    # fixed  effects design matrix construction
              kronecker(diag(p), matrix(1 ,nrow = m, ncol = 1)))                # fixed  effects design matrix construction
X           = X_t[,-2]                                                          # fixed  effects design matrix for one cell culture strain 
Z           = kronecker(diag(b), matrix(rep(1,k), nrow = k))                    # random effects design matrix
beta        = matrix(c(5,0,0,2), nrow = p)                                      # fixed effects null hypothesis parameter
sig_eps     = .5                                                                # error variance
sig_b       = 1                                                                 # random effects variance
mu_y        = X %*% beta                                                        # marginal expectation parameter
Sigma_y     = sig_b*(Z %*% t(Z)) + sig_eps*diag(n)                              # marginal covariance parameter
y           = mvrnorm(1, mu_y, Sigma_y)                                         # data sampling
T           = rep(paste0("T", 1:4), each = m)                                   # treatment factor
E           = rep(paste0("E", 1:b), each = k)                                   # experimental unit factor
D           = data.frame(TRM = T, EU = E, y = y)                                # data formatting
write.csv(D, "./_data/anova-nested-subsampling-design-data-set.csv", row.names = FALSE)   
```

```{r, eval = F}
# visualization as "stripchart"
library(latex2exp)
pdf(                                                                            
file        = "./_figures/anova-nested-subsampling-design-data.pdf",                                           
width       = 10,                                                                     
height      = 6) 
par(                                                                    
family      = "sans",                                                  
mfcol       = c(1,1),                                                  
pty         = "m",                                                     
bty         = "l",                                                     
lwd         = 1,                                                       
las         = 2,                                                       
mgp         = c(2,1,0),                                                
xaxs        = "i",                                                   
yaxs        = "i",                                                  
font.main   = 1,                                                       
cex         = 1.2,
cex.main    = 1.4,
xpd         = FALSE)
xpos        = 1:b
xjit        = rep(xpos, each = k) + runif(n, -0.15, 0.15)
tcols       = c("T1"="gray20","T2"="gray40","T3"="gray60","T4"="gray80")
ptcol       = adjustcolor(tcols[D$TRM], alpha.f = 0.9)
plot(
xjit, 
D$y, 
pch         = 19, 
col         = ptcol,
xlim        = c(.5,12.5),
ylim        =  c(1,10),
xaxt        = "n", 
xlab        = " ",
ylab        = "Observation", 
main        = "Nested subsampling design data")
axis(
1, 
at          = unique(xpos), 
labels      = rep(paste0("E", 1:b)),
las         = 1)
abline(
v           = c(3.5, 6.5, 9.5), 
lty         = 3, 
col         = "grey70")
mtext(
c("T1","T2","T3","T4"), 
side        = 1, 
line        = 3,
at          = c(2,5,8,11), 
cex         = 1.4, 
las         = 1)
dev.off()
```


![Nested subsampling design data example.](./_figures/anova-nested-subsampling-design-data.pdf){#fig-anova-nested-subsampling-design-data fig-align="center" width=70%} 

As for the randomized block design, it can be shown that the presence of random terms such as the $\mu_{ik}$s induces 
correlations in the marginal error and  data covariance matrices. In the case of 
nested subsampling designs, these correlations induce an increase of the between
sum of squares such that, when not accounted for, on average higher $F$-values 
are observed than predicted based on independent observational unit error terms. 
@fig-anova-nested-subsampling-design-f-simulation\unskip B shows this shift of observed $F$-values 
under the null hypothesis when computed based on samples from @eq-anova-nested-subsampling 
as opposed to computed based on samples from @eq-anova-lm as shown in 
@fig-anova-nested-subsampling-design-f-simulation\unskip A. Again, as shown 
in @fig-anova-nested-subsampling-design-f-simulation\unskip C, using a linear mixed 
model that properly accounts for the induced error correlations renders the analysis 
less prone to false positive results. 

```{r, echo = F, eval = F}
library(nlme)                                                                   # linear mixed models
library(MASS)                                                                   # multivariate normal distribution
set.seed(2)                                                                     # reproducibility 
p           = 4                                                                 # number of experimental groups 
m           = 30                                                                # number of observational units per group  
a           = 3                                                                 # number of animals per group 
b           = p*a                                                               # number of animals in total
k           = m/a                                                               # number of observation units per animal
n           = p*m                                                               # number of observational units in total
X_t         = cbind(                                                            # fixed  effects design matrix construction
              matrix(1, nrow = n, ncol = 1),                                    # fixed  effects design matrix construction
              kronecker(diag(p), matrix(1 ,nrow = m, ncol = 1)))                # fixed  effects design matrix construction
X           = X_t[,-2]                                                          # fixed  effects design matrix for one cell culture strain 
Z           = kronecker(diag(b), matrix(rep(1,k), nrow = k))                    # random effects design matrix
beta        = matrix(c(5,0,0,0), nrow = p)                                      # fixed effects null hypothesis parameter
sig_eps     = 1                                                                 # error variance
sig_b       = 1                                                                 # random effects variance
nsamp       = 10000                                                             # number of samples per model
I_n         = diag(n)                                                           # error covariance matrix structure      
F_lm        = rep(NaN, nsamp)                                                   # F-values for linear model
F_lmm       = rep(NaN, nsamp)                                                   # F-values for linear mixed model
F_lme       = rep(NaN, nsamp)                                                   # F-values for linear mixed model estimated using LME

# linear model simulations
mu_y        = X %*% beta                                                        # marginal expectation parameter
Sigma_y     = sig_eps*I_n                                                       # marginal covariance parameter
for(i in 1:nsamp){                                                              # sample iterations
  y         = mvrnorm(1, mu_y, Sigma_y)                                         # data sampling
  T         = rep(paste0("T", 1:4), each = m)                                   # treatment factor
  E         = rep(paste0("E", 1:b), each = k)                                   # experimental unit factor
  D         = data.frame(TRM = T, EU = E, y = y)                                # data formatting
  R         = lm(y ~ TRM, data = D)                                             # ANOVA 
  F_lm[i]   = summary(R)$fstatistic[1]                                          # F-value
}

# linear mixed model simulations of a nested subsampling design
mu_y        = X %*% beta                                                        # marginal expectation parameter
Sigma_y     = sig_b*(Z %*% t(Z)) + sig_eps*I_n                                  # marginal covariance parameter
for(i in 1:nsamp){                                                              # sample iterations
  y         = mvrnorm(1, mu_y, Sigma_y)                                         # data sampling
  T         = rep(paste0("T", 1:4), each = m)                                   # treatment factor
  E         = rep(paste0("E", 1:b), each = k)                                   # experimental unit factor
  D         = data.frame(TRM = T, EU = E, y = y)                                # data formatting
  R         = lm(y ~ TRM, data = D)                                             # ANOVA     
  F_lmm[i]  = summary(R)$fstatistic[1]                                          # F-value appropriate linear model analysis
  F_lme[i]  = anova(lme(y ~ TRM, data = D, random = ~ 1|EU))["TRM", "F-value"]  # F-value appropriate linear mixed model analysis
}
```

```{r eval = F}
library(latex2exp)
pdf(                                                                            
file        = "./_figures/anova-nested-subsampling-design-f-simulation.pdf",                                           
width       = 15,                                                                     
height      =  5)                                                                                                                                         
par(                                                                    
family      = "sans",                                                  
mfcol       = c(1,3),                                                  
pty         = "m",                                                     
bty         = "l",                                                     
lwd         = 1,                                                       
las         = 1,                                                       
mgp         = c(2,1,0),                                                
xaxs        = "i",                                                   
yaxs        = "i",                                                  
font.main   = 1,                                                       
cex         = 1.2,
cex.main    = 1)

# linear model with linear model analysis
f_min       = 0                                                                  
f_max       = 6                                                                 
f_res       = 1e3                                                                
f           = seq(f_min,f_max, len=f_res)                                     
p_F         = df(f,p-1,n-p)                                                          
hist(
F_lm,                                                                       
breaks      = 100,                                                                
col         = "gray90",                                                           
prob        = TRUE,                                                               
xlim        = c(f_min, f_max),                                                    
ylim        = c(0,.8),                                                            
xlab        = "F-value",                                                                
ylab        = "",                                                                 
main        = "Independent observational units \n Linear model analysis")
lines(
f,                                                                          
p_F,                                                                       
lwd         = 2,                                                                  
col         = "darkorange")      

# linear mixed model with linear model analysis
f_min       = 0                                                                  
f_max       = 6                                                               
f_res       = 1e3                                                                
f           = seq(f_min,f_max, len=f_res)                                     
p_F         = df(f,p-1,n-p) 
hist(
F_lmm,                                                                       
breaks      = 100,                                                                
col         = "gray90",                                                           
prob        = TRUE,                                                               
xlim        = c(f_min, f_max),                                                    
ylim        = c(0,1.2),                                                            
xlab        = "F-value",                                                                
ylab        = "",                                                                 
main        = "Nested subsampling design \n Linear model analysis")
lines(
f,                                                                          
p_F,                                                                       
lwd   = 2,                                                                  
col   = "darkorange")      

# linear mixed model with linear model analysis
numDF       = anova(lme(y ~ TRM, data = D, random = ~ 1|EU))["TRM","numDF"]     # DF approximation
denDF       = anova(lme(y ~ TRM, data = D, random = ~ 1|EU))["TRM","denDF"]     # DF approximation
f_min       = 0                                                                  
f_max       = 6                                                                
f_res       = 1e3                                                                
f           = seq(f_min,f_max, len=f_res)                                     
p_F         = df(f,numDF,denDF) 
hist(
F_lme,                                                                       
breaks      = 100,                                                                
col         = "gray90",                                                           
prob        = TRUE,                                                               
xlim        = c(f_min, f_max),                                                    
ylim        = c(0,.8),                                                            
xlab        = "F-value",                                                                
ylab        = "",                                                                 
main        = "Nested subsampling design \n Linear mixed model analysis")
lines(
f,                                                                          
p_F,                                                                       
lwd   = 2,                                                                  
col   = "darkorange")      
dev.off()                                                       
```

![Increase of false positives in a nested subsampling design when analyzed using
a linear model ANOVA, but not a linear mixed model ANOVA.](./_figures/anova-nested-subsampling-design-f-simulation.pdf){#fig-anova-nested-subsampling-design-f-simulation fig-align="center" width=100%}

# Bibliographic remarks {#sec-bibliographic-remarks}

@fisher1925a and @scheffe1959 provide classical treatments of ANOVA as a 
computational procedure.  @winer1971 and @maxwell2018 provide excellent, largely 
matrix calculus-free introductions to ANOVA as a linear model. @hays1994 is a good
text for newcomers to statistics. @snijders2012 provide a general non-technical 
introduction to linear mixed models and @yu2022 provides an introduction to linear mixed models
in the context of cell and animal research.

# Supplementary Material {#sec-supplementary-material}

\footnotesize

## The sum symbol {#sec-the-sum-symbol}

The sum symbol is defined as
\begin{equation}
\sum_{i=1}^{n} x_i = x_1 + x_{2} + \cdots + x_{n}.
\end{equation}
Here:

* $\Sigma$ denotes the Greek letter Sigma, mnemonic for Sum,
* the subscript $i = 1$ specifies the running index and the starting index,
* the superscript $n$ specifies the ending index, and
* $x_1, x_2, \dots, x_n$ are the summands.

For meaningful use of the summation symbol, it is essential to specify the lower 
and upper limits of summation via the subscript and superscript.The particular 
choice of symbol for the running index is irrelevant to the value of the sum; indeed,
\begin{equation}
\sum_{i=1}^n x_i = \sum_{j=1}^n x_j.
\end{equation}
Sometimes the running index is given as an element of an index set.
For example, if the index set $I := {1,5,7}$ is defined, then
\begin{equation}
\sum_{i \in I}x_i := x_1 + x_5 + x_7.
\end{equation}
In the following, we present several examples illustrating the use of the summation symbol.

*Summation of predefined summands.* Let $x_1 := 2$, $x_2 := 10$, and $x_3 := -4$. Then
\begin{equation}
\sum_{i=1}^3 x_i = x_1 + x_2 + x_3 = 2 + 10 - 4 = 8.
\end{equation}
*Summation of weighted predefined summands.* Let $x_1 := 2$, $x_2 := 10$, and $x_3 := -4$.
Furthermore, let the weighting coefficients be
$a_1 := \frac{1}{2}$, $a_2 := \frac{1}{5}$, and $a_3 := 2$.
Then
\begin{equation}
\sum_{i=1}^3 a_i x_i = a_1x_1 + a_2x_2 + a_3x_3 = \frac{1}{2}\cdot 2 + \frac{1}{5}\cdot 10 + 2\cdot (-4) = 1 + 2 - 8 = -5.
\end{equation}
Expressions of the form $\sum_{i=1}^n a_i x_i$ are referred to as *linear combinations* 
of $x_1,\dots,x_n$ with coefficients or weights $a_1,\dots,a_n$.

*Summation of natural numbers*.
\begin{equation}
\sum_{i=1}^5 i = 1 + 2 + 3 + 4 + 5 = 15.
\end{equation}

*Summation of even natural numbers.*
\begin{equation}
\sum_{i=1}^5 2i = 2\cdot 1 + 2\cdot 2 + 2\cdot 3 + 2\cdot 4 + 2\cdot 5 = 2 + 4 + 6 + 8 + 10 = 30.
\end{equation}

*Summation of odd natural numbers.*
\begin{equation}
\sum_{i=1}^5 (2i - 1) = 2\cdot 1 - 1 + 2\cdot 2 - 1 + 2\cdot 3 - 1 + 2\cdot 4 - 1 + 2\cdot 5 - 1 = 1 + 3 + 5 + 7 + 9 = 25.
\end{equation}

*Nested sums* The computational procedure for evaluating nested sums, i.e. expressions involving
multiple sequential sum signs is best understood by expanding it from the inside 
outward, noting that during each iteration of the inner sum, the index of the 
outer sum remains fixed, The following example illustrates this:
\begin{align}
\begin{split}
\sum_{i=1}^2 \sum_{j = 1}^3 (x_i + y_j) 
& = \sum_{i=1}^2 (x_i + y_1 + x_i + y_2 + x_i + y_3) \\
& = (x_1 + y_1 + x_1 + y_2 + x_1 + y_3) + (x_2 + y_1 + x_2 + y_2 + x_2 + y_3). 
\end{split}
\end{align}

## Total sum-of-squares partition {#sec-total-sum-of-squares-partition}

By definition of the $\mbox{SQT}$, $\mbox{SQB}$ and $\mbox{SQW}$, we have
\begin{align}
\begin{split}
\mbox{SQT} 
& = \sum_{i=1}^{p} \sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}\right)^{2} \\
& = \sum_{i=1}^{p} \sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}+\bar{y}_{i}-\bar{y}\right)^{2} \\
& = \sum_{i=1}^{p} \sum_{j=1}^{n_{i}}\left(\left(y_{ij}-\bar{y}_{i}\right)+\left(\bar{y}_{i}-\bar{y}\right)\right)^{2} \\
& = \sum_{i=1}^{p} \sum_{j=1}^{n_{i}}\left(\left(y_{ij}-\bar{y}_{i}\right)^{2}+2\left(y_{ij}-\bar{y}_{i}\right)\left(\bar{y}_{i}-\bar{y}\right)+\left(\bar{y}_{i}-\bar{y}\right)^{2}\right) \\
& =\sum_{i=1}^{p}\left(\sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}+\sum_{j=1}^{n_{i}} 2\left(y_{ij}-\bar{y}_{i}\right)\left(\bar{y}_{i}-\bar{y}\right)+\sum_{j=1}^{n_{i}}\left(\bar{y}_{i}-\bar{y}\right)^{2}\right)
\\
& =\sum_{i=1}^{p}\left(\sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}+2\left(\bar{y}_{i}-\bar{y}\right) \sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)+n_{i}\left(\bar{y}_{i}-\bar{y}\right)^{2}\right) \\
& =\sum_{i=1}^{p}\left(\sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}+2\left(\bar{y}_{i}-\bar{y}\right) \sum_{j=1}^{n_{i}}\left(y_{ij}-\frac{1}{n_{i}} \sum_{j=1}^{n_{i}} y_{ij}\right)+n_{i}\left(\bar{y}_{i}-\bar{y}\right)^{2}\right) \\
& =\sum_{i=1}^{p}\left(\sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}+2\left(\bar{y}_{i}-\bar{y}\right)\left(\sum_{j=1}^{n_{i}} y_{ij}-\sum_{j=1}^{n_{i}}\left(\frac{1}{n_{i}} \sum_{j=1}^{n_{i}} y_{ij}\right)\right)+n_{i}\left(\bar{y}_{i}-\bar{y}\right)^{2}\right) \\
& =\sum_{i=1}^{p}\left(\sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}+2\left(\bar{y}_{i}-\bar{y}\right)\left(\sum_{j=1}^{n_{i}} y_{ij}-\frac{n_{i}}{n_{i}} \sum_{j=1}^{n_{i}} y_{ij}\right)+n_{i}\left(\bar{y}_{i}-\bar{y}\right)^{2}\right)
\\ 
& =\sum_{i=1}^{p}\left(\sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}+2\left(\bar{y}_{i}-\bar{y}\right)\left(\sum_{j=1}^{n_{i}} y_{ij}-\sum_{j=1}^{n_{i}} y_{ij}\right)+n_{i}\left(\bar{y}_{i}-\bar{y}\right)^{2}\right) \\
& =\sum_{i=1}^{p}\left(\sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}+n_{i}\left(\bar{y}_{i}-\bar{y}\right)^{2}\right) \\
& =\sum_{i=1}^{p} \sum_{j=1}^{n_{i}}\left(y_{ij}-\bar{y}_{i}\right)^{2}+\sum_{i=1}^{p} n_{i}\left(\bar{y}_{i}-\bar{y}\right)^{2} \\
& =\mbox{SQW}+\mbox{SQB}
\end{split}
\end{align} 

## The normal distribution {#sec-the-normal-distribution}

The normal (or Gaussian) distribution is one of the most commonly encountered
mathematical models for random phenomena, such as measurement errors in scientific 
experiments. A key reason for its importance is the *Central Limit Theorem*, 
which states that the sum of many independent random variables - regardless of 
their individual distributions - tends to converge to a normal distribution. Thus, 
the idea that, beyond the explanations offered by a deterministic scientific theory,
numerous unknown factors influence any given observation provides strong motivation 
for modelling error variables with the normal distribution. Formally, the normal
distribution is described by the probability density function.
$$
p : \mathbb{R} \to \mathbb{R}_{>0}, x \mapsto p(x) := \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{1}{2\sigma^2}(x-\mu)^2\right)
$$ {#eq-normal-disribution}
The expectation parameter $\mu$ determines the location of the function on the
$x$ axis, the variance parameter $\sigma^2$ the width of the curve. In general, 
a probability density function defined on the real numbers represents 
probabilities through densities. To determine the probability that a normally 
distributed random variable falls within a given interval, one integrates its 
probability density function over that interval in the Riemannian sense. Since 
interval length is measured uniformly along the real line, regions of high probability 
density - such as those near the expectation  parameter $\mu$ - correspond to 
higher probabilities of the random variable taking values there. The situation is analogous
to the concept of densities in physics: only multiplying a non-zero volume (interval) 
by the density (probability) of some material yields a non-zero mass (probybility mass). 
Repeated sampling with replaced of a random variable (or simultaneous sampling of 
many identical copies of a single random variable) produces data that, when analyzed 
statistically, reflect the characteristic distribution of the underlying mathematical model.

```{r, eval = F, echo = F}
library(latex2exp)
set.seed(1)
pdf(                                                                            
file        = "./_figures/anova-normal-distribution.pdf",                                           
width       = 11,                                                                     
height      = 4) 
par(
family      = "sans",
mfcol       = c(1,3),
pty         = "s",
bty         = "l",
lwd         = 1,
las         = 1,
mgp         = c(2,1,0),
xaxs        = "i",
yaxs        = "i",
font.main   = 1,
cex         = 1,
cex.main    = 1)
x_min       = -4                                            
x_max       = 4                                            
x_res       = 1e3                                          
x           = seq(x_min, x_max, length.out = x_res)        
mu          = 0                                            
sigsqr      = 1                                         

# Wahrscheinlichkeitstheorie
plot(
x,
dnorm(x,mu,sqrt(sigsqr)),
type        = "l",
col         = "black",
lwd         = 1,
main        = "Probability density function",
xlab        = "x",
ylab        = "",
ylim        = c(0,.4))

# Daten
n           = 100
D           = rnorm(n,mu,sqrt(sigsqr))
plot(
D,
rep(0,n),
type        = "p",
lwd         = 1,
main        = "Realizations",
xlab        = "x",
ylab        = "",
xlim        = c(-4,4),
ylim        = c(-2,2),
pch         = 21,
col         = "white",
bg          = "gray60",
cex         = 1)

# Deskriptive Statistik
hist(
D,
main        = "Descriptive statistics",
xlim        = c(-4,4),
ylab        = "Counts",
xlab        = "x",
ylim        = c(0,25))
dev.off()
```

![The normal distribution from a probability theory-, data-, and descriptive statistics- 
perspective.](./_figures/anova-normal-distribution.pdf){#fig-anova-normal-distribution fig-align="center" width=100%}


# References {.unnumbered}
\small

